Estimating duration in the presence of misclassification

### 2024-01-28 transmission.tex
to dos

\section{Data} \label{transmission:sec:data}

@@

I use the period from the expansion of the CIS, starting 31st August 2020, until there was significant vaccination rollout\todo{insert date}.

@@

\todo[inline]{
    I should discuss the literature on estimating incidence based on SARS-CoV-2 prevalence data.
    In particular: \textcites{mccabeCISincidence}{abbottCISincidence}. Also Toffe.
    However, I think this goes better in the introduction, framing the whole thesis because it is important in understanding the motivation for the whole work.
    At the same place, I could review the literature on mechanistic models using REACT and CIS.
}

@@

The logit transformation serves two purposes.
First, it ensures that the incidence proportion or prevalence is between 0 and 1.
Second, at low values of incidence or prevalence, linear growth of $\logit(\pi_i)$ implies approximately exponential in $\pi_i$~\autocite[see discussion in][and references therein]{ealesAppropriately}.
Approximately exponential growth is what mechanistic models predict, and penalising towards exponential growth has previously been shown to improve estimates of prevalence by balancing over and under-fitting~\autocite{ealesAppropriately}.

@@

\todo{Insert constant incidence assumption methodology}

@@

The final step is poststratifying the incidence proportion to the CIS target population and subpopulations of interest.
I draw 500\todo{check number of samples} samples from the approximate posterior distribution estimated.
For each posterior draw and day, the incidence proportion in the (sub)population is the weighted average of the incidence proportion in each stratum, where the weights are the proportion of the population in each stratum.
The population numbers for each stratum are provided by the Office for National Statistics.

@@

I take information on the contact rates, $\kappa_{aa'}(t)$, from several data sources, using pre-existing methodology~\autocites{vanleeuwenTime}{vanleeuwenAugmenting}.
%; these estimates have been used for previous modelling of the COVID-19 pandemic~\autocites{vanleeuwenTime}{birrellRealtime}.
% These start with estimates of contact in each setting, based on \textcite{vanleeuwenAugmenting}.
This method takes pre-pandemic baseline estimates of: $\kappa_{aa'}$ from the POLYMOD study~\autocite{mossongSocial}, and time use from the UK Time Use Survey (UKTUS)~\autocite{UKTUS}.
POLYMOD asked a representative sample of the UK population about the contacts they make each day, including information on each contact's age and location (home, work, school, leisure, transport, or other).
UKTUS asked a representative sample of the UK population about how much time they spent doing various activities (\eg shopping and spending time at home) each day.
\Textcite{vanleeuwenAugmenting} combined these two datasets to estimate the rate of contacts between age groups during each activity.
\Textcite{birrellRealtime} then modify these contact rates based on measured changes in time use during the pandemic.
These changes are measured using Google mobility data~\autocite{googleCOVID19} and school attendance data~(unpublished).
Google mobility data is based on the location of Android users over time, relative to a pre-pandemic baseline.
The mobility and school data are averaged over a week then used to down or up-weight the number of contacts in each setting, before combining contacts across settings to form $\kappa_{aa'}(t)$.
$\kappa_{aa'}(t)$ is assumed to be piecewise constant, changing only at the start of each week.
For further details see \textcite{vanleeuwenAugmenting} and \textcite[supplementary material]{birrellRealtime}.

@@

The transmission model represents the infection process (introduced in \cref{E-inc-prev:sec:infection-process}).
A variety of different approaches can be taken to link a transmission model to observations.
In the context of CIS, this requires representing the prevalence and observation processes (introduced in \cref{E-inc-prev:sec:prevalence-process,E-inc-prev:sec:observation-process} respectively).
The method to do so is known as the observation model.
The formulation of this model is heavily dependent on the data available.
This means that there is a wide variety of models used.
This section  explains the model I propose for fitting to the CIS prevalence data.

@@

\Textcite{osogamiClosed} define $P$ as well-matching an arbitrary distribution $G$ if $P$ and $G$ have the same first three moments.
They then prove that EC distributions have the following properties:
\begin{itemize}
    \item If there exists a phase-type distribution $P'$ that well-matches $G$ then there exists an EC distribution $P$ that well-matches $G$.
    \item The parameters of $P$ are available in closed form, which \textcite{osogamiClosed} give.
    \item The number of phases used by the $P$ calculated by these expressions is at most one more than the minimum across all well-matching $P'$.
\end{itemize}

>>>>

\todo[inline]{
    Lit review to go in introduction chapter, framing the whole thesis because it is important in understanding the motivation for the whole work.
    Would include \textcites{mccabeCISincidence}{abbottCISincidence} on backcalculation. Also Toffe?
    As well as the literature on mechanistic models using REACT and CIS I have previously circulated.
}
\section{Data} \label{transmission:sec:data}

@@

I use the period from the expansion of the CIS, starting Mon 31st August 2020 until Sun 24th Jan 2021 (inclusive).
This period is chosen because there were very few positives in CIS in summer 2020 due to low prevalence and a smaller sample size, and ends when vaccination crossed 10\% of England's population.

@@

@@

The logit transformation serves two purposes.
First, it ensures that the incidence proportion or prevalence is between 0 and 1.
Second, at low values of incidence or prevalence, linear growth of $\logit(\pi_i)$ implies approximately exponential in $\pi_i$~\autocite[see discussion in][and references therein]{ealesAppropriately}.
Penalising towards exponential growth is a natural formulation because it is what mechanistic models predict over short periods of time.
Furthermore, it has previously been shown to improve estimates of prevalence by balancing over and under-fitting~\autocite{ealesAppropriately}.

@@

For $\vec{P} = \matr{S} \vec{I}$ to have a unique solution, $\vec{I}$ and $\vec{P}$ need the same dimension.
However, I am using a period of prevalence which starts after the first infections.
These prior infections will still contribute to the prevalence.
I assume that the incidence up to and including the first day of prevalence data, at $t = 1$, is constant at $I_0$.
Therefore, $P_1 = I_0 \sum_{i=1}^{\dmax} S(i)$ and $I_0 = P_1 / \sum_{i=1}^{\dmax} S(i)$.
For day $t$, the contribution of the incidence for $t < 1$ can be calculated as $P'_t = I_0 \sum_{i=t+1}^{\dmax} S(i)$.
Subtracting $\vec{P'}$ from $\vec{P}$ before applying the forward solving procedure accounts for the contribution of the incidence prior to the first day of prevalence data.

@@

The final step is poststratifying the incidence proportion to the CIS target population and subpopulations of interest.
For this, I first draw 500 samples from the approximate posterior distribution estimated.
For each posterior draw and day, the incidence proportion in the (sub)population is the weighted average of the incidence proportion in each stratum, where the weights are the proportion of the population in each stratum.
The population numbers for each stratum are provided by the Office for National Statistics.

@@

I take information on the contact rates, $\kappa_{aa'}(t)$, from several data sources, using the methodology of \textcites{vanleeuwenTime}{vanleeuwenAugmenting}.
%; these estimates have been used for previous modelling of the COVID-19 pandemic~\autocites{vanleeuwenTime}{birrellRealtime}.
% These start with estimates of contact in each setting, based on \textcite{vanleeuwenAugmenting}.
This method takes pre-pandemic baseline estimates of: $\kappa_{aa'}$ from the POLYMOD study~\autocite{mossongSocial}, and time use from the UK Time Use Survey (UKTUS)~\autocite{UKTUS}.
POLYMOD asked a representative sample of the UK population about the contacts they make each day, including information on each contact's age and location (home, work, school, leisure, transport, or other).
UKTUS asked a representative sample of the UK population about how much time they spent doing various activities (\eg shopping and spending time at home) each day.
\Textcite{vanleeuwenAugmenting} combined these two datasets to estimate the rate of contacts between age groups during each activity.
\Textcite{birrellRealtime} then modify these contact rates based on measured changes in time use during the pandemic.
These changes are measured using Google mobility data~\autocite{googleCOVID19} and school attendance data~(unpublished).
Google mobility data is based on the location of Android users over time, relative to a pre-pandemic baseline.
The mobility and school data are averaged over a week then used to down or up-weight the number of contacts in each setting, before combining contacts across settings to form $\kappa_{aa'}(t)$.
$\kappa_{aa'}(t)$ is assumed to be piecewise constant, changing only at the start of each week.
For further details see \textcite{vanleeuwenAugmenting} and \textcite[supplementary material]{birrellRealtime}.

@@

The transmission model represents the infection process (introduced in \cref{E-inc-prev:sec:infection-process}).
A variety of different approaches can be taken to link a transmission model to observations.
In the context of CIS, this requires representing the prevalence and observation processes (introduced in \cref{E-inc-prev:sec:prevalence-process,E-inc-prev:sec:observation-process} respectively).
The method to do so is known as the observation model.
The formulation of this model is heavily dependent on the data available.
This means that there is a wide variety of models used.
This section  explains the novel modelling approach I propose for fitting to the CIS prevalence data.

@@

\Textcite{osogamiClosed} define a phase-type distribution $P$ as well representing an arbitrary distribution $G$ if $P$ and $G$ have the same first three moments.
They then prove that EC distributions have the following properties.
\begin{itemize}
    \item Let $\set{W}$ be the set of all phase-types distributions that well represent $G$ and $w$ be the minimum number of phases of any member of $\set{W}$. If $\set{W}$ is non-empty, then there exists a $l$-phase EC distribution $P' \in \set{W}$.
    \item Furthermore, at least one such $P'$ has $l \leq w + 1$. Fewer phases is generally beneficial as it reduces the time and space complexity of working with the distribution.
    \item The parameters of a $P'$ with $l \leq w + 1$ can be found using a closed-form expression (given by \textcite{osogamiClosed}). This is beneficial because finding arbitrary members of $\set{W}$ is a hard problem.
\end{itemize}
